Online Multiclass Linear Classification with Bandit Feedback

### 2018-11-27 arxiv-version/main.tex
Almost finish properties of Chebyshev polynomials

% Math symbols and commands
\newcommand{\R}{\mathbb{R}}  % set of real numbers
\newcommand{\indicator}[1]{\mathbf{1}\left[#1 \right]} % indicator
\newcommand{\ip}[2]{\left\langle #1, #2 \right\rangle} % inner product
\newcommand{\norm}[1]{\left\| #1 \right\|}  % norm of a vector or a matrix
%\DeclareMathOperator*{\deg}{deg}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\Exp}{\mathbf{E}}  % expected value

@@

In this section, we show how construct a mapping $\phi$ of points in the unit
ball $B_d$ of $\R^d$ into a high dimensional inner product space that has the
property if any set of labeled examples in $B_d$ is linearly separable with a
margin $\gamma$, applying the mapping $\phi$ makes the examples \emph{strongly}
linearly separable with a margin $\gamma'$ and their norms are bounded by $R'$.
The parameters $\gamma'$ and $R'$ are functions of the old margin $\gamma$ and
the number of classes $K$, and the are specified in the theorems below.

@@

The idea behind all the mappings is that, on a compact set, multivariate
polynomials uniformly approximate any continuous function. This is result is a
special case of the well known Stone-Weierstrass theorem~\citep[Section
10.10]{Davidson-Donsig-2010}. In our case, we use a multivariate polynomials to
approximate the indicator function corresponding to the intersection of $K-1$
halfspaces. This polynomial separates examples in one class from examples in the
other classes. To be able to quantify $\gamma'$ and $R'$ as well as time
complexity of evaluating $k(x,x')$ we need to quantify certain parameters
of the approximating polynomial (degree, norm, range of values).
The parameters are quantified in \autoref{theorem:polynomial-approximation-1}
and \autoref{theorem:polynomial-approximation-2}
stated below.

@@

\begin{theorem}[Polynomial approximation of intersection of halfspaces (1)]
\label{theorem:polynomial-approximation-1}
Let $w_1, w_2, \dots, w_K \in \R^d$ be such that $\sum_{i=1}^K \norm{w_1}^2 \le 1$.
Let $\gamma > 0$. There exists a polynomial $p:\R^d \to \R$ such that
\begin{itemize}
\item $p(x) \ge 1$ for all $x \in \bigcap_{i=2}^K \{ x \in \R^d ~:~ \ip{w_1}{x} \ge \ip{w_i}{x} + \gamma \}$
\item $p(x) \le -1$ for all $x \in \bigcup_{i=2}^K \{ x \in \R^d ~:~ \ip{w_1}{x} \le \ip{w_i}{x} - \gamma \}$
\item $\deg(p) \le ???$
\item $\norm{p} \le ???$
\item $|p(x)| \le ???$ for all $x \in \R^d$ such that $\norm{x} \le 1$
\end{itemize}
\end{theorem}

\begin{theorem}[Polynomial approximation of intersection of halfspaces (2)]
\label{theorem:polynomial-approximation-2}
Let $w_1, w_2, \dots, w_K \in \R^d$ be such that $\sum_{i=1}^K \norm{w_1}^2 \le 1$.
Let $\gamma > 0$. There exists a polynomial $p:\R^d \to \R$ such that
\begin{itemize}
\item $p(x) \ge 1$ for all $x \in \bigcap_{i=2}^K \{ x \in \R^d ~:~ \ip{w_1}{x} \ge \ip{w_i}{x} + \gamma \}$
\item $p(x) \le -1$ for all $x \in \bigcup_{i=2}^K \{ x \in \R^d ~:~ \ip{w_1}{x} \le \ip{w_i}{x} - \gamma \}$
\item $\deg(p) \le ???$
\item $\norm{p} \le ???$
\item $|p(x)| \le ???$ for all $x \in \R^d$ such that $\norm{x} \le 1$
\end{itemize}
\end{theorem}

@@

To construct the polynomial $p$ we use Chebyshev polynomials of the first kind.
Chebyshev polynomials of the fist kind form an infinite sequence of polynomials
$T_0(z), T_1(z), T_2(z), \dots$ of single real variable $z$. They are defined
by recurrence
\begin{align*}
T_0(z) & = 1 \; , \\
T_1(z) & = z \; , \\
T_{n+1}(z) & = 2zT_n(z) - T_{n-1}(z) & \text{for $n \ge 1$.}
\end{align*}

\begin{proposition}
Chebyshev polynomials satisfy for any $n \ge 0$
\begin{enumerate}
\item $|T_n(z)| \le 1$ for all $z \in [-1,1]$.
\item If $z > 1$ then $T_n(z) \ge 1 + n^2(z - 1)$.
\item Absolute value of all coefficients of $T_n(z)$ is at most $2^n$.
\end{enumerate}
\end{proposition}

>>>>

% Math symbols and commands
\newcommand{\R}{\mathbb{R}}  % set of real numbers
\newcommand{\indicator}[1]{\mathbf{1}\left[#1 \right]} % indicator
\newcommand{\ip}[2]{\left\langle #1, #2 \right\rangle} % inner product
\newcommand{\norm}[1]{\left\| #1 \right\|}  % norm of a vector or a matrix
%\DeclareMathOperator*{\deg}{deg}
\DeclareMathOperator*{\argmax}{argmax}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\Exp}{\mathbf{E}}  % expected value
\DeclareMathOperator{\diff}{d \!} % differential

@@

In this section, we show how construct a mapping $\phi$ of points in the unit
ball of $\R^d$ into a high dimensional inner product space that has the
property if any set of labeled examples in the unit ball is linearly separable with a
margin $\gamma$, applying the mapping $\phi$ makes the examples \emph{strongly}
linearly separable with a margin $\gamma'$ and their norms are bounded by $R'$.
The parameters $\gamma'$ and $R'$ are functions of the old margin $\gamma$ and
the number of classes $K$, and the are specified in the theorems below.

@@

The idea behind all the mappings is that, on a compact set, multivariate
polynomials uniformly approximate any continuous function. This result is a
special case of the well known Stone-Weierstrass theorem; see
e.g.~\citep[Section~10.10]{Davidson-Donsig-2010}. In our case, we use
multivariate polynomials to approximate the indicator function corresponding to
the intersection of $K-1$ halfspaces. This polynomial separates examples in one
class from examples in the other classes. To be able to quantify $\gamma'$ and
$R'$ as well as time complexity of evaluating $k(x,x')$ we need to quantify
certain parameters of the approximating polynomial (degree, norm, range of
values). The parameters are quantified in
\autoref{theorem:polynomial-approximation-1} and
\autoref{theorem:polynomial-approximation-2} stated below.

@@

\begin{theorem}[Polynomial approximation of intersection of halfspaces I]
\label{theorem:polynomial-approximation-1}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1}, \norm{v_2}, \dots, \norm{v_m} \le 1$.
Let $\gamma > 0$. There exists a polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $p(x) \ge 1$ for all $x \in \bigcap_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \ge \gamma \right\}$
\item $p(x) \le -1$ for all $x \in \bigcup_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \le - \gamma \right\}$
\item $\deg(p) \le ???$
\item $\norm{p} \le ???$
\item $|p(x)| \le ???$ for all $x \in \R^d$ such that $\norm{x} \le 1$
\end{enumerate}
\end{theorem}

\begin{theorem}[Polynomial approximation of intersection of halfspaces II]
\label{theorem:polynomial-approximation-2}
Let $v_1, v_2, \dots, v_m \in \R^d$ be vectors such that $\norm{v_1}, \norm{v_2}, \dots, \norm{v_m} \le 1$.
Let $\gamma > 0$. There exists a polynomial $p:\R^d \to \R$ such that
\begin{enumerate}
\item $p(x) \ge 1/2$ for all $x \in \bigcap_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \ge \gamma \right\}$
\item $p(x) \le -1/2$ for all $x \in \bigcup_{i=1}^m \left\{ x \in \R^d ~:~ \norm{x} \le 1, \ \ip{v_i}{x} \le - \gamma \right\}$
\item $\deg(p) \le ???$
\item $\norm{p} \le ???$
\item $|p(x)| \le ???$ for all $x \in \R^d$ such that $\norm{x} \le 1$
\end{enumerate}
\end{theorem}

@@

To construct the polynomial $p$ we use Chebyshev polynomials of the first kind.
Chebyshev polynomials of the fist kind form an infinite sequence of polynomials
$T_0(z), T_1(z), T_2(z), \dots$ of single real variable $z$. They are defined
by the recurrence
\begin{align*}
T_0(z) & = 1 \; , \\
T_1(z) & = z \; , \\
T_{n+1}(z) & = 2zT_n(z) - T_{n-1}(z) & \text{for $n \ge 1$.}
\end{align*}
Chebyshev polynomials have a lot of interesting properties.
We will need properties listed in
\autoref{proposition:properties-of-chebyshev-polynomials} below.
Interested reader can learn more about Chebyshev polynomials
from the book by \cite{Mason-Handscomb-2002}.

\begin{proposition}[Properties of Chebyshev polynomials]
\label{proposition:properties-of-chebyshev-polynomials}
Chebyshev polynomials satisfy
\begin{enumerate}
\item $\deg(T_n) = n$ for all $n \ge 0$.
\item If $n \ge 1$, the leading coefficient of $T_n(z)$ is $2^{n-1}$.
\item $|T_n(z)| \le 1$ for all $z \in [-1,1]$ and all $n \ge 0$.
\item $T_n(z) \ge 1 + n^2(z - 1)$ for all $z \ge 1$ and all $n \ge 0$.
\item All coefficients of $T_n(z)$ lie in the interval $\left[-(1+\sqrt{2})^n, (1+\sqrt{2})^n \right]$.
\end{enumerate}
\end{proposition}

\begin{proof}[Proof of \autoref{proposition:properties-of-chebyshev-polynomials}]
The first two proprties easily follow from the recurrence.

The third property is a direct consequence of
\begin{equation}
\label{equation:trigonometric-chebyshev}
T_n(\cos(\theta)) = \cos(n \theta) \qquad \quad \text{for all $\theta \in \R$.}
\end{equation}
since for any $x \in [-1,1]$ there exists $\theta \in \R$ such that $\cos(\theta) = x$.
Equation \eqref{equation:trigonometric-chebyshev} is proved by induction on $n$. Indeed, by definition
$$
T_0(\cos(\theta)) = 1 = \cos(0 \theta) \qquad \text{and} \qquad T_1(\cos(\theta)) = \cos(\theta) \; .
$$
For $n \ge 1$, we have
\begin{align*}
T_{n+1}(\cos(\theta))
& = 2 \cos(\theta) T_n(\cos(\theta)) - T_{n-1}(\cos(\theta)) \\
& = 2 \cos(\theta) \cos(n \theta) - \cos((n-1)\theta)) \; ,
\end{align*}
where the last step follow by induction hypothesis.
It remains to show that the last expression equals $\cos((n+1)\theta)$.
This can be derived from the trigonometric formula
$$
\cos(\alpha \pm \beta) = \cos(\alpha) \cos(\beta) \mp \sin(\alpha) \sin(\beta) \; .
$$
By substituting $\alpha = n \theta$ and $\beta = \theta$, we get two equations
\begin{align*}
\cos((n+1) \theta) & = \cos(n \theta) \cos(\theta) - \sin(n \theta) \sin(\theta) \; , \\
\cos((n-1) \theta) & = \cos(n \theta) \cos(\theta) + \sin(n \theta) \sin(\theta) \; .
\end{align*}
Summing them yields
$$
\cos((n+1)\theta) + \cos((n-1) \theta) = 2 \cos(n \theta) \cos(\theta)
$$
which finishes the proof of equation \eqref{equation:trigonometric-chebyshev}.

To verify the fourth property we first show that
\begin{equation}
\label{equation:hyperbolic-chebyshev}
T_n(\cosh(\theta)) = \cosh(n \theta) \qquad \quad \text{for all $\theta \in \R$.}
\end{equation}
The proof of equation~\eqref{equation:hyperbolic-chebyshev} is the same
as the proof of equation \eqref{equation:trigonometric-chebyshev} except
that we replace $\cos$ and $\sin$ with $\cosh$ and $\sinh$ respectively.

Note that $\cosh(\theta) = \frac{e^{\theta} + e^{-\theta}}{2}$ is an even
continuous function that maps $\R$ onto $[1,+\infty)$, is strictly decreasing
on $(-\infty,0]$, and is strictly increasing on $[0,\infty)$. Thus the fourth
property is equivalent to
$$
T_n(\cosh(\theta)) \ge 1 + n^2 (\cosh(\theta) - 1) \qquad \text{for any $\theta \ge 0$.}
$$
For $\theta = 0$, according to \eqref{equation:hyperbolic-chebyshev} both sides
are equal to $1$. The proof will be finished, if we prove that the derivative of
the left hand side is greater or equal to the derivative of the right hand side.
That is, we need to prove that
\begin{equation}
\label{equation:hyperbolic-chebyshev-derivative-inequality}
\frac{\diff T_n(\cosh(\theta))}{\diff \theta} \ge \frac{\diff (1 + n^2 (\cosh(\theta) - 1))}{\diff \theta} \qquad \text{for any $\theta \ge 0$.}
\end{equation}
Using \eqref{equation:hyperbolic-chebyshev} and recalling that $\diff \cosh(\theta) / \diff \theta = \sinh(\theta)$ we
can express the derivatives as
\begin{align*}
\frac{\diff T_n(\cosh(\theta))}{\diff \theta} & = \frac{\diff \cosh(n \theta))}{\diff \theta} = n \sinh(n \theta) \; , \\
\frac{\diff (1 + n^2 (\cosh(\theta) - 1))}{\diff \theta} & = n^2 \sinh(\theta) \; .
\end{align*}
TODO

To verify the fifth property let $a_n$ be the magnitude of the largest
coefficient of $T_n(z)$. The recurrence for Chebyshev polynomials
implies that the sequence $a_0, a_1, a_2, \dots$ satisfies
\begin{align*}
a_0 & = a_1 = 1 \; , \\
a_{n+1} & \le 2 a_n + a_{n-1} & \text{for $n \ge 1$.}
\end{align*}
It remains to show that $a_n \le (1 + \sqrt{2})^n$ for all $n \ge 0$. We do this
by induction on $n$. For $n=0$ and $n=1$ the inequality trivially holds. For $n
\ge 1$, we have
\begin{align*}
a_{n+1} \le 2 a_n + a_{n-1}
& \le 2 (1 + \sqrt{2})^n + (1 + \sqrt{2})^{n-1} \\
& = (1 + \sqrt{2})^{n-1} (2 (1 + \sqrt{2}) + 1) \\
& = (1 + \sqrt{2})^{n-1} (3 + 2\sqrt{2}) \\
& = (1 + \sqrt{2})^{n-1} (1 + \sqrt{2})^2 \\
& = (1 + \sqrt{2})^{n+1} \; .
\end{align*}

\end{proof}

Let $r = \left\lceil \log_2(2m) \right\rceil$ and $s = \left\lceil \sqrt{\frac{1}{\gamma}} \right\rceil$.
We define the polynomial $p:\R^d \to \R$ as
$$
p(x) = m + \frac{1}{2} - \sum_{i=1}^m \left( T_s(1 - \ip{v_i}{x}) \right)^r \; .
$$
It remains to show that $p$ has properties 1--5.
